---
title: "`comrad` demo"
author: "Théo Pannetier"
output: rmarkdown::html_vignette
date: "4/3/2020"
vignette: >
  %\VignetteIndexEntry{demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(comrad); library(ape)
```

## Introduction and biological context
`comrad` is an individual-based simulation of an evolving community of 
organisms. At every (discrete, non-overlapping) generation, individuals sample a
number of offspring in a Poisson with parameter their fitness value. Offspring 
undergo mutation and (when the criterion is met) ecological speciation, 
then replace the parent community. 

Competition between individuals causes selection to be disruptive, and over 
time, individuals form discrete clusters in the (unidimensional) trait space, 
i.e., species. Individuals in a cluster are assigned as a new species whenver 
their (trait) distance to the nearest individual outside the cluster is 
$\Delta z \ge 0.1$.

Over time, the lineages formed by these clusters branch again and again, until
the community occupies the entire trait space. The community is then at 
equilibrium, and fitness is around 1 everywhere on the trait space. I'm 
interested in reconstructing the phylogeny of the community, and observing how
the rates of speciation, extinction, and trait evolution change with the density
of species in the community.

## Model details
Fitness $W$ depends on the trait value of the focal individual ($z_i$), and the 
trait values of all other individuals in the community.

$$
  W(z_i) = e^{r(1 - \frac{N_{eff}(z_i)}{K(z_i)})}
$$
where $r$ is the baseline growth rate, $N_{eff}(z_i)$ the effective population 
size experienced by the individual, which is the sum of competition terms 
$\alpha(z_i,z_j)$:
$$
  N_{eff}(z_i) = \sum_{j=1}^n \alpha(z_i,z_j)
$$
where
$$
  \alpha(z_i,z_j) = e^{-\frac{(z_i - z_j)^2}{2\sigma_{\alpha}^2} }
$$
varies from 0 to 1; $\sigma_{\alpha}$ is the SD for the competition kernel,
and called `competition_sd` in `comrad`.

$K(z_i)$ is the carrying capacity at $z_i$, i.e. the
number of individuals that can maintain the same trait value *and* $W \ge 1$.
$$
  K(z_i) = K_{opt} * e^{-\frac{(z_i - z_{opt})^2}{2\sigma_{K}^2}}
$$
$K_{opt}$ is the carrying capacity at the optimal trait value $z_{opt}$, set at
1000 and 0 respectively. $\sigma_{K}$ is the SD for the "niche" kernel, and is
dubbed `carrying_cap_sd` in `comrad`.

## Installation
```{r lib_comrad, eval=FALSE}
remotes::install_github("TheoPannetier/comrad")
```

## Running the simulation
To run the simulation, call `run_simulation()`.
```{r run_sim}
temp_path_to_output <- paste0(tempfile("comrad_temp_output"), ".csv")

comrad_tbl <- comrad::run_simulation(
  path_to_output = temp_path_to_output,
  nb_gens = 1000
)
```

Default values for the parameters of the model are:
* $\sigma_K$ = 0.5
* $\sigma_\alpha$ = 0.2
* $z_{opt}$ = 0
* $K_{opt}$ = 1000
* $r$ = 1

Default values of the parameters can be called with `comrad::default_X()` where 
X is replaced with the name of the parameter.

The community starts with 10 individuals with trait 0, of species `"#89ae8a"`.
I use colour hexes for species names, this makes for some nice and consistent
plotting.

`path_to_output` is the only argument without a default. Providing the name of a 
`.csv` file will save the community at every sampled generation in that file.
Saved data contains one row per individual, and contains that individual's trait
value $z$, its species label, and the label of its ancestor species (to build
the phylogeny). Because the sie of the community is quite large with default
parameter values (about 2,500 individuals per generation at equilibrium for default values), only 
half of the individuals (randomly sampled) in the community are saved – 
`sampling_prop = 0.5`. I found this is sufficient to have enough info on the
community.
Generations are sampled every twice a tenth of the order of magnitude of `nb_gens`, i.e. every 20 for `nb_gens = 100`, every 200 for
`nb_gens = 1000`, every 200 again for `nb_gens = 3000`, etc.

`path_to_output = NULL` will return the output of the last generation (only) 
instead of saving it.

Saved communities can be loaded from the csv with `readr::read_csv()` as usual,
but I also save some lines of metadata in a header. For an easy loading,
use `comrad::read_comrad_tbl()` instead.

```{r read_comm}
comrad_tbl <- comrad::read_comrad_tbl(path_to_file = temp_path_to_output)
comrad_tbl
```

## Data viz
The distribution and evolution of the community in the trait space can be 
visualized with `plot_comm_trait_evolution()`

```{r data_viz}
comrad_tbl %>% comrad::plot_comm_trait_evolution(xgrain = 10, ygrain = 0.01)
```

Hexes of a colour represent a single species. Not much has happened by 
generation 1000, but you should see a first speciation event. Note the slight
protraction between the actual branching time and the time speciation is
detected.

Trait distribution at a single generation can also be plotted:
```{r data_viz2}
comrad_tbl %>% comrad::plot_comm_traits(generation = 1000)
```

A third angle to visualize the community is through the growth curve:
```{r data_viz3}
comrad_tbl %>% plot_comm_size()
```

The community initially grows exponentially, until the carrying 
capacity at $z = 0$ is reached. The number of individuals will keep growing more
slowly, as clusters push each other away and explore new parts of the trait 
space.

The simulation can take quite long, but interesting things happen after many 
generations. Below is an example with $\sigma_z = 0.1$ & $\sigma_K = 1$.

```{r load_sim, echo=FALSE, warning=FALSE}
comrad_tbl <- comrad::read_comrad_tbl("demo_data/comrad_output_example.csv", skip = 18)
```

```{r data_viz4}
comrad_tbl %>% comrad::plot_comm_trait_evolution(xgrain = 200, ygrain = 0.05)
comrad_tbl %>% comrad::plot_comm_traits(generation = max(.$t))
comrad_tbl %>% plot_comm_size()
```

Note that only half of the individuals were sampled in the output, so the size
of the community at the end is about 10,000, not 5,000.

## Phylogeny reconstruction
From the simulation ouput, build the phylogeny with `sim_to_phylo()`.
```{r phylo_build}
phylo <- comrad_tbl %>% comrad::sim_to_phylo()
phylo
```

For visualization, it's `ape` as usual:

```{r phylo_plot}
phylo %>% ape::plot.phylo()
phylo %>% ape::ltt.plot()
```

Work in progress :)
