---
title: "Simulate diversification in a community with `comrad`"
author: "Théo Pannetier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction
Welcome to `comrad`!

`comrad` is an individual-based simulation of a community that undergoes 
adaptation, speciation, and extinction in a simple ecological setting. 
The model is based on the work of Pontarp et al. (2012) and other similar
eco-evo models.

`comrad` hasn't found its definitive purpose: it's work-in-progress, and bound
to evolve, branch and adapt in the near future. Below is a showcase of what it
can do already.

```{r setup}
library(comrad)
```

# The model
Develop a section about the model, including equations.

# Simulation

# Results

## Overview
Population growth, trait distribution, species

## Phylogenies

# References
Pontarp, M., J. Ripa, and P. Lundberg, 2012. On the origin of phylogenetic structure in competitive metacommunities. Evolutionary Ecology Research 14:269–284.
